---
title: "D-value"
author: "Freddy Hernandez and Karen Rivera"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{D value}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##"
)
```

# P-value for example 1 of Demidenko (2016)

In this section we will replicate the example 1 of Demidenko (2016) using the function `t_test` of the package 'stests' as shown below.

```{r, message=FALSE}
require(stests)

t_test(meanx=249, varx=20^2, nx=10000,
       meany=250, vary=20^2, ny=10000,
       alternative='less', mu=0, var.equal=TRUE)
```

The results obtained by the function `t_test` are the same obtained by Demidenko

# How does the sample size $n$ of each group influence the P-value?

In the following figure the P-value is calculated for many values of $n$, in which it is evident that the P-value decreases and then in $n=2000$, this parameter is below the level of usual significance of 5%

```{r, echo=FALSE}
p1 <- function(n) {
  t_test(meanx=249, varx=20^2, nx=n,
         meany=250, vary=20^2, ny=n,
         alternative='less', mu=0, var.equal=TRUE)$p.value
}
p1 <- Vectorize(p1)
n <- seq(from=5, to=10000, by=100)
plot(x=n, y=p1(n), pch=20, ylab='P-value', las=1, ylim=c(0, 0.5))
abline(h=0.05, lty='dashed', col='tomato')
text(x=4000, y=0.07, '0.05', col='tomato')

```

# P-value for example 2 of Demidenko (2016)

In this section we will replicate the example 2 of Demidenko (2016) using the function `t_test` of the package 'stests' as shown below.

```{r, message=FALSE}
t_test(meanx=15, varx=6^2, nx=7,
       meany=10, vary=6^2, ny=7,
       alternative='greater', mu=0, var.equal=TRUE)
```

We see that the statistics coincide with what is obtained in the article, the value of P within the function `t_test` is calculated with a $t$-student distribution whereas in the article it is obtained through the normal distribution, on the other hand, the size of the sample is small ($n=6$) because of this is the difference between the methods.

# How does the sample size of each group influence the P-value?

In the following figure the P-value is calculated for many values of $n$, in which it is evident that the P-value decreases and then in $n=9$, this parameter is below the level of usual significance of 5%.

```{r, echo=FALSE}
p2 <- function(n) {
  t_test(meanx=15, varx=6^2, nx=n,
         meany=10, vary=6^2, ny=n,
         alternative='greater', mu=0, var.equal=TRUE)$p.value
}
p2 <- Vectorize(p2)
n <- seq(from=7, to=20, by=1)
plot(x=n, y=p2(n), pch=20, ylab='P-value', las=1, ylim=c(0, 0.1))
abline(h=0.05, lty='dashed', col='tomato')
text(x=14, y=0.06, '0.05', col='tomato')
```

# Function to calculate the D-value

Hypothesis testing:

- The null hypothesis is :

$$H_0: \mu_1 = \mu_2$$


- The alternative hypothesis can be:

    $$H_A: \mu_1 < \mu_2$$

    $$H_A: \mu_1 \neq \mu_2$$

    $$H_A: \mu_1 > \mu_2$$

The function `d_meantest` presented below is to calculate the D-value by simulation for the case of comparison of two means from normal populations.

```{r}
require(stests)
args(d_meantest)
```

# D-value for example 1 of Demidenko (2016)

We will replicate the example 1 of Demidenko (2016) using the function `d_meantest`.

```{r}
 d_meantest(meanx=250,meany=249, varx=20^2, vary=20^2, alternative='less')
```

# D-value for the example 2 of Demidenko (2016)

We will replicate the example  2 of Demidenko (2016) using the function `d_meantest`.

```{r}
d_meantest(meanx=10, meany=15, varx =6^2, vary=6^2, alternative='less')
```

# How to change the D-value for the different values of $\bar{x}$ y $\bar{y}$?

To answer this question we will take the data from example 1 of Demidenko (2016), the value of $x = 249$ will be fixed while the value of $\bar{y}$ will be moving away from $\bar{x}$. Recall that the value of $\bar{y}$ represents the average weight of patients who consumed the drug to lose weight. The value of the sample desviation $s$ for each group will remain constant at 20 lbs.

The following figure shows the evolution of the D-value as the weight changes $\bar{y}$ for the group of people who did take the medication.

```{r, echo=FALSE}
d_mtest1 <- function(meanx, varx, meany, vary,
        alternative='less', nrep=1000000) {
  x <- rnorm(n=nrep, mean=meanx, sd=sqrt(varx))
  y <- rnorm(n=nrep, mean=meany, sd=sqrt(vary))
  if (alternative == 'less')
    dvalue <-  mean(y > x)
  if (alternative == 'greater')
    dvalue <-  mean(y < x)
  if (alternative == 'two.sided') {
    dvalue1 <-  mean(y > x)
    dvalue2 <-  mean(y < x)
    dvalue <- 2 * min(c(dvalue1, dvalue2))
  }
  return(dvalue)
}
means <- 320:180
aux_function <- function(x)
  d_mtest1(meanx=250, meany=x, varx=20^2, vary=20^2, alternative='less', nrep=100000)
aux_function <- Vectorize(aux_function)

deltas <- aux_function(means)

plot(x=means, y=deltas, type='b', las=1, pch=19, cex=0.4,
     xlab=expression(bar(y)), ylab='D-value')
abline(v=250, lty='dashed', col='green4')
text(x=269, y=0.04, col='green4',
     'Sample mean of patients in the placebo group')
```

# References
- Demidenko, E. (2016). The p-Value You Can't Buy. The American Statistician, 70(1): 33-38.
