---
title: "Tests to compare covariance matrices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tests_Sigmas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(stests)
```

# Introduction

Let $\mathbf{X}_{1i}, \mathbf{X}_{2i}, \ldots, \mathbf{X}_{N_i i}$ be a random sample from the population 
$N_p(\boldsymbol{\mu}_g, \boldsymbol{\Sigma}_g)$ indexed by $i = 1,2,\ldots,g$. 
In other words, we have $g$ samples as follows

\[
\begin{matrix}
\mathbf{X}_{11}, \mathbf{X}_{21}, \ldots, \mathbf{X}_{N_1 1} \\
\mathbf{X}_{12}, \mathbf{X}_{22}, \ldots, \mathbf{X}_{N_2 2} \\
\vdots \\
\mathbf{X}_{1g}, \mathbf{X}_{2g}, \ldots, \mathbf{X}_{N_g g}
\end{matrix}
\]

It is not necessary to have equal $N_i$. The main objective in this vignette is to use test to study the following hypothesis.

\[
H_{0}: \Sigma_{1} = \Sigma_{2} = \cdots = \Sigma_{g} = \Sigma
\]

\[
H_{A}: \text{at least one } \Sigma \text{ is different}
\]

# Box-M test


