---
title: "Tests to compare covariance matrices"
output: rmarkdown::html_vignette
bibliography: ref_multiv_test.bib
vignette: >
  %\VignetteIndexEntry{Tests_Sigmas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, message=FALSE}
library(stests)
```

# Introduction

Let $\mathbf{X}_{1i}, \mathbf{X}_{2i}, \ldots, \mathbf{X}_{N_i i}$ be a random sample from the population 
$N_p(\boldsymbol{\mu}_g, \boldsymbol{\Sigma}_g)$ indexed by $i = 1,2,\ldots,g$. 
In other words, we have $g$ samples as follows

$$
\begin{matrix}
\mathbf{X}_{11}, \mathbf{X}_{21}, \ldots, \mathbf{X}_{N_1 1} \\
\mathbf{X}_{12}, \mathbf{X}_{22}, \ldots, \mathbf{X}_{N_2 2} \\
\vdots \\
\mathbf{X}_{1g}, \mathbf{X}_{2g}, \ldots, \mathbf{X}_{N_g g}
\end{matrix}
$$

It is not necessary to have equal $N_i$. The main objective in this vignette is to use test to study the following hypothesis.

$$
H_{0}: \Sigma_{1} = \Sigma_{2} = \cdots = \Sigma_{g} = \Sigma
$$

$$
H_{A}: \text{at least one } \Sigma \text{ is different}
$$

# Box-M test
@box1949 proposed this test and the statistic test $\varphi$ is given by:

$$
\varphi = -2 \rho \log(\lambda)
$$

Under true $H_{0}$, the statistic 

$$
\varphi \sim \chi^{2}_{p(p+1)(g-1)/2}
$$

Where $\rho$, $\log(\lambda)$ and $S$ are obtained by as

$$
\rho = 1 - \frac{2p^{2} + 3p - 1}{6(p + 1)(g - 1)}
\left( \sum_{i=1}^{g} \frac{1}{n_i} - \frac{1}{n} \right)
$$

$$
\log(\lambda) = 
\frac{n \, \log(|S|) - \sum_{i=1}^{g} n_i \, \log(|S_i|)}{-2}
$$


$$
S = \frac{1}{n} \sum_{i=1}^{g} n_i S_i
$$

with 

$$
n_i = N_i - 1
$$

$$
n = n_1 + n_2 + \ldots n_g
$$


This test seems to be good if each $N_i$ exceeds 20, and if $g$ and $p$ do not exceed 5 (@Mardia_et_1992, page 140).

# Bartlett's test or modified LRT
xxx proposed this test and the statistic is given by:

$$
M = n \log(|S|) - \sum_{i=1}^{g} n_i \log(|S_i|)
$$

Under true $H_{0}$, the statistic 

$$
M \sim \chi^{2}_{p(p+1)(g-1)/2}
$$

The matrix $S$ and $n$ are the same as in the Box-M test.

Note: @schott2007 claims that since the sample covariance matrix $S_i$ is singular if $n_i < p$, this likelihood ratio test is valid only if $n_i \ge p$ for $i = 1,2, \ldots, g$.

# Wald Schott test
@schott2001 (page 27) proposed this test and the statistic is given by:

$$
W = \frac{n}{2} \left\{ 
\sum_{i=1}^{g} \frac{n_i}{n} \, \mathrm{tr}(S_i S^{-1} S_i S^{-1})
- 
\sum_{i=1}^{g} \sum_{j=1}^{g} \frac{n_i n_j}{n^{2}} \, \mathrm{tr}(S_i S^{-1} S_j S^{-1})
\right\}.
$$

Under true $H_{0}$, the statistic 

$$
W \sim \chi^{2}_{p(p+1)(g-1)/2}
$$

The matrix $S$ and $n$ are the same as in the Box-M test.

# References

